Declarative Deployment Reconfiguration of Pulverised Collective Adaptive Systems at Runtime \thanks{ This work has been supported by the Italian PRIN Project COMMON-WEARS (2020HCWWLP) and the EU/MUR FSE REACT-EU PON R\&I 2014-2020. }

### 2023-05-20 paper-2023-acsos-pulverization.tex
Pass on bib
Towards clarification of II.C on middlewares/runtimes and dsls -- what do you think @danysk

\acrodef{dsl}[DSL]{domain-specific language}
\acrodef{iot}[IoT]{Internet of Things}
\acrodef{cas}[CAS]{collective adaptive system}
\acrodef{qos}[QoS]{quality of service}
\acrodef{rssi}[RSSI]{received signal strength indicator}
\acrodef{poi}[PoI]{point of interest}

@@

The \emph{deployment view} is a well-known architectural viewpoint for software systems,
concerned with the mapping of \emph{software components} to \emph{physical machines}~\cite{DBLP:journals/software/Kruchten95}, 
 and supported by modelling notations like \emph{UML Deployment diagrams}.
%
Here, we provide a brief introduction to deployment and reconfiguration, based on the conceptual characterisation of~\cite{DBLP:journals/jss/ArcangeliBL15,carzaniga1998characterization}.
%
A \emph{site} is a set of computers (\emph{hosts}) that may host a \emph{software system}, i.e., a product described by a coherent collection of artefacts and consisting of a set of \emph{deployment units} that can be independently operated.
%
\emph{(Software) Deployment} is the process of moving and making a software system available and operational from one or more \emph{producer sites} to a target set of \emph{consumer sites}, also called the \emph{deployment domain}.
%
The \emph{deployment plan} defines 
 the mapping between the software system
 and the deployment domain,
 possibly augmented with further information (e.g., metadata, constraints, preferences).
%
Common deployment-related activities 
 include: (i) \emph{release/update} of the software system at the producer site;
 (ii) \emph{installation/deinstallation} at/from the consumer sites;
 (iii) \emph{activation/deactivation}, for starting/stopping the components;
 (iv) \emph{reorganisation} of the software system;
 and (v) \emph{redistribution}, i.e., changing the deployment plan.
%
According to the analytical framework of~\cite{DBLP:journals/jss/ArcangeliBL15},
  the problem of (automatic) deployment of distributed software system can be addressed by considering
 (i) the nature of the software to be deployed (e.g., how the software system is split into components);
 (ii) the nature of the deployment domain (i.e. the characteristics and topology of the available infrastructure);
 (iii) how the deployment is designed (e.g., how the deployment plan is specified);
 and 
 (iv) how the deployment is performed (e.g., how deployment activities are carried out).
%
By an operational point of view, 
 deployment may be supported by a \emph{middleware}~\cite{DBLP:journals/cacm/GazisK22}.
%
Related work covering these issues is provided in \Cref{sec:rw}.
%
In the following, we introduce the deployment approach 
  that we extend and upon which we develop \ourframework{}.

@@

As mentioned in \Cref{sec:background:dep},
 the deployment and re-configuration of a software system
 may be supported -- following the well-known layered architectural style -- by a \emph{middleware} or \emph{runtime}~\cite{DBLP:journals/cacm/GazisK22}
 encapsulating, among others, deployment and reconfiguration services.

In principle, there are several possibilities to implement a pulverisation configurator and runtime:
as a library
configured by a program written in a general-purpose language,
as a framework/runtime configured by means of configuration files,
or as a \ac{dsl} configured with its own custom syntax.
%
These approaches have different trade-offs.
%
A (well designed) library can be flexible as the host language can be used to express peculiar configurations
and easier to adopt to users acquainted with the host language;
but its configuration may quickly become imperative and harder to maintain as complexity grows.
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but its flexibility is limited,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
Finally, a \ac{dsl} can be designed to be as expressive as needed,
but it has high maintenance cost (as the language maintenance costs stack upon those of the library/API) and, potentially,
a steep learning curve due to the need of learning a new (custom) language.
%
Recent evolution in programming languages,
however,
open to an additional strategy:
the construction of \emph{internal} \acp{dsl},
hybrids between libraries and stand-alone (external) \acp{dsl}.
%
In fact, modern languages such as Kotlin, Groovy, Ruby, and Scala~\cite{Riti2018}
provide specific syntactic features to enable the construction of APIs whose ergonomy is akin to the one of a dedicated language,
but that are valid fragments in the host language.
%
Although internal \acp{dsl} are de facto libraries in the host programming language
(there is no clear boundary defining when a library becomes a \ac{dsl}\footnote{\url{https://archive.is/wip/xAeiX}}),
from a practical perspective
they allow for great flexibility and ergonomy
(although not total as the one of a custom \ac{dsl}, as they are subject to the syntactic constraints of the host language)
while retaining a reduced maintenance cost
(as the host language ecosystem and tooling can be reused directly)
and a gentle learning curve compared to stand-alone \acp{dsl}
(as the syntax will be largely familiar to the host language users).
%
The killer feature of internal \acp{dsl}, compared to external ones,
is the counterpass of the limited syntax imposed by the host:
if some peculiar entity or process cannot be express by the \ac{dsl},
the designer can fall back to the host language directly,
with no need to resort to interoperability features,
and with the possibility to exploit all the libraries designed for the host language.
%
The internal DSL approach has thus been successfully applied in several domains,
spanning from
build systems\footnote{\url{https://archive.is/5xtaN}}
and ontologies~\cite{DBLP:journals/jossw/Balhoff16} to
hardware design~\cite{DBLP:journals/trets/SerreP20} to
logic~\cite{2pkt-swx16} and aggregate programming~\cite{DBLP:journals/softx/CasadeiVAP22}.
%

@@

\bibliographystyle{IEEEtran}
\bibliography{bibliography}

>>>>

\acrodef{api}[API]{application-program interface}
\acrodef{dsl}[DSL]{domain-specific language}
\acrodef{iot}[IoT]{Internet of Things}
\acrodef{cas}[CAS]{collective adaptive system}
\acrodef{qos}[QoS]{quality of service}
\acrodef{rssi}[RSSI]{received signal strength indicator}
\acrodef{poi}[PoI]{point of interest}

@@

The \emph{deployment view} is a well-known architectural viewpoint for software systems,
concerned with the mapping of \emph{software components} to \emph{physical machines}~\cite{DBLP:journals/software/Kruchten95}, 
 and supported by modelling notations like \emph{UML Deployment diagrams}.
%
Here, we provide a brief introduction to deployment and reconfiguration, based on the conceptual characterisation of~\cite{DBLP:journals/jss/ArcangeliBL15,carzaniga1998characterization}.
%
A \emph{site} is a set of computers (\emph{hosts}) that may host a \emph{software system}, i.e., a product described by a coherent collection of artefacts and consisting of a set of \emph{deployment units} that can be independently operated.
%
\emph{(Software) Deployment} is the process of moving and making a software system available and operational from one or more \emph{producer sites} to a target set of \emph{consumer sites}, also called the \emph{deployment domain}.
%
The \emph{deployment plan} defines 
 the mapping between the software system
 and the deployment domain,
 possibly augmented with further information (e.g., metadata, constraints, preferences).
%
Common deployment-related activities 
 include: (i) \emph{release/update} of the software system at the producer site;
 (ii) \emph{installation/deinstallation} at/from the consumer sites;
 (iii) \emph{activation/deactivation}, for starting/stopping the components;
 (iv) \emph{reorganisation} of the software system;
 and (v) \emph{redistribution}, i.e., changing the deployment plan.
%
According to the analytical framework of~\cite{DBLP:journals/jss/ArcangeliBL15},
  the problem of (automatic) deployment of distributed software system can be addressed by considering
 (i) the nature of the software to be deployed (e.g., how the software system is split into components);
 (ii) the nature of the deployment domain (i.e. the characteristics and topology of the available infrastructure);
 (iii) how the deployment is designed (e.g., how the deployment plan is specified);
 and 
 (iv) how the deployment is performed (e.g., how deployment activities are carried out).
%
By an operational point of view, 
 deployment may be supported by a so-called \emph{runtime} or \emph{middleware}~\cite{DBLP:journals/cacm/GazisK22}.
%
Related work covering these issues is provided in \Cref{sec:rw}.
%
In the following, we introduce the deployment approach 
  that we extend and upon which we develop \ourframework{}.

@@

As mentioned in \Cref{sec:background:dep},
 the deployment and re-configuration of a software system
 may be supported -- following the well-known layered architectural style -- by a \emph{runtime} or \emph{middleware}~\cite{DBLP:journals/cacm/GazisK22}
 encapsulating, among others, deployment and reconfiguration services.
%
The runtime could be reified into a deployment unit \emph{per se},
 or can be implemented as part of the deployment units of the software system to be deployed---e.g., taking the form of an \emph{\ac{api}} or a \emph{framework}.
%
In order to take decisions about deployment and reconfiguration, 
 such a runtime has to be \emph{configured}
 for the software application at hand,
 i.e., it has to be given a \emph{deployment plan} capturing deployment mappings, constraints, and policies.
%

In principle, there are several possibilities to implement a deployment plan:
%
(i) declaratively, through a set of \emph{configuration files};
(ii) through a \emph{\ac{api}/library}, implemented in a general-purpose language;
or 
(iii) through a \ac{dsl}~\cite{dsl-book-voelter}. % configured with its own custom syntax.
%
These approaches have different trade-offs.
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but its flexibility is limited,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
A (well designed) library can be flexible as the host language can be used to express peculiar configurations
and easier to adopt to users acquainted with the host language;
but its configuration may quickly become imperative and harder to maintain as complexity grows.
%
Finally, a \ac{dsl}~\cite{dsl-book-voelter} can be designed to be as expressive as needed,
but it has high maintenance cost (as the language maintenance costs stack upon those of the library/\ac{api}) and, potentially,
a steep learning curve due to the need of learning a new (custom) language.
%
Recent evolution in programming languages,
however,
open to an additional strategy:
the construction of \emph{internal} \acp{dsl},
hybrids between libraries and stand-alone (external) \acp{dsl}~\cite{dsl-book-voelter}.
%
In fact, modern languages such as Kotlin, Groovy, Ruby, and Scala~\cite{Riti2018}
provide specific syntactic features to enable the construction of APIs whose ergonomy is akin to the one of a dedicated language,
but that are valid fragments in the host language.
%
Although internal \acp{dsl} are de facto libraries in the host programming language
(there is no clear boundary defining when a library becomes a \ac{dsl}\footnote{\url{https://archive.is/wip/xAeiX}}),
from a practical perspective
they allow for great flexibility and ergonomy
(although not total as the one of a custom \ac{dsl}, as they are subject to the syntactic constraints of the host language)
while retaining a reduced maintenance cost
(as the host language ecosystem and tooling can be reused directly)
and a gentle learning curve compared to stand-alone \acp{dsl}
(as the syntax will be largely familiar to the host language users).
%
The killer feature of internal \acp{dsl}, compared to external ones,
is the counterpass of the limited syntax imposed by the host:
if some peculiar entity or process cannot be express by the \ac{dsl},
the designer can fall back to the host language directly,
with no need to resort to interoperability features,
and with the possibility to exploit all the libraries designed for the host language.
%
The internal DSL approach has thus been successfully applied in several domains,
spanning from
build systems\footnote{\url{https://archive.is/5xtaN}}
and ontologies~\cite{DBLP:journals/jossw/Balhoff16} to
hardware design~\cite{DBLP:journals/trets/SerreP20} to
logic~\cite{2pkt-swx16} and aggregate programming~\cite{DBLP:journals/softx/CasadeiVAP22}.
%

@@

\bibliographystyle{IEEEtranDOI}
\bibliography{bibliography}
